---
title: "Welling Univariate Modeling Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Univariate Assignment

Read in tree data, metadata can be found in: `./data/tree_metadata.txt`

```{r}
trees <- read.csv('treedata_subset.csv', header=T)

head(trees)
```

1. Carry out an exploratory analysis using the tree dataset. 
Develop and compare models for species cover for a habitat generalist
[_*Acer rubrum*_ (Red maple)](http://www.durhamtownship.com/blog-archives/pix/November1407.jpg) 
and a habitat specialist [_*Abies fraseri*_ (Frasier fir)](https://upload.wikimedia.org/wikipedia/commons/d/d0/Abies_fraseri_Mitchell.jpg). 
Because this dataset includes both continuous and discrete explanatory
variables use the function `Anova` in the packages `car` as such.
This will estimate partial effect sizes, variance explained, and p-values for 
each explanatory variable included in the model.

```{r}
library(car)

#subseting the data 
A_rubrum <- trees[trees$species=="Acer rubrum",]


#Exploratory analysis for A. rubrum:
Ar_mod1 <- aov(A_rubrum$cover~A_rubrum$elev+A_rubrum$tci+A_rubrum$streamdist+
                 A_rubrum$disturb+A_rubrum$beers)
summary(Ar_mod1)
Anova(Ar_mod1, type=3)

#based on initial model, disturbance is not significant and will be 
# taken out of model
Ar_mod2 <- aov(A_rubrum$cover~A_rubrum$elev+A_rubrum$tci+A_rubrum$streamdist+
                 A_rubrum$beers)
summary(Ar_mod2)
Anova(Ar_mod2, type=3)

#tci will be taken out of the model as well based on significance values
#of previous model
Ar_mod3 <- aov(A_rubrum$cover~A_rubrum$elev+A_rubrum$streamdist+
                 A_rubrum$beers)
summary(Ar_mod3)
Anova(Ar_mod3, type=3)

AIC(Ar_mod1)
AIC(Ar_mod2)
AIC(Ar_mod3)
```
For Acer rubrum, it seems as though elevation, distance from streams, tci, and beers
are important for the cover of this species.

```{r}
#subsetting the data
A_fraseri <- trees[trees$species=="Abies fraseri",]

#Exploratory analysis for A. fraseri:
Af_mod1 <- aov(A_fraseri$cover~A_fraseri$elev+A_fraseri$tci+A_fraseri$streamdist+
                 A_fraseri$disturb+A_fraseri$beers)
summary(Af_mod1)
Anova(Af_mod1, type=3)

#based on the first model, it will only be run with elevation
Af_mod2 <- aov(A_fraseri$cover~A_fraseri$elev)
summary(Af_mod2)
Anova(Af_mod2, type=3)

AIC(Af_mod1)
AIC(Af_mod2)

```
Elevation seems to be the only important aspect for Abies fraseri cover.


 

Compare the p-values you observe using the function `Anova` to those generated
 using `summary`. 

For each species address the following additional questions:

        * how well does the exploratory model appear to explain cover?
        * which explanatory variables are the most important?
        * do model diagnostics indicate any problems with violations of
        OLS assumptions?
        * are you able to explain variance in one species better than another, 
        why might this be the case?

2. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

```{r}
#acer_glm = glm(cover ~ tci + elev + ... , data= my_data, 
#             family='poisson')
```

For assessing the degree of variation explained you can use a 
pseudo-R-squared statistic (note this is just one of many possible)

```{r}
#pseudo_r2 = function(glm_mod) {
#    1 -  glm_mod$deviance / glm_mod$null.deviance
 #}
```

Compare the residual sums of squares between the traditional OLS 
 and glm models using `anova` (Note: not `Anova`) as such

```{r}
#anova(acer_ols, acer_glm)
```

Does it appear that changing the error distribution changed the
results much? In what ways? 

3. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?

4. (optional) Examine the behavior of the function `step()` using the 
exploratory models developed above. This is a very simple and not very
robust machine learning stepwise algorithm that uses AIC to select a 
best model. By default it does a backward selection routine. 

5. (optional) Develop a model for the number of species in each site 
(i.e., unique plotID). This variable will also be discrete so the Poisson
may be a good starting approximation. Side note: the Poisson
distribution converges asymptotically on the Gaussian distribution as the 
mean of the distribution increases. Thus Poisson regression does not differ
much from traditional OLS when means are large. 