---
title: "Welling_Spatial_Statistics_Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial Modeling Assignment

```{r}
library(vegan)
library(nlme)
library(vegan)
data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

1) Examine if there is evidence of spatial dependence in a rare and a common
species in the BCI tree dataset


```{r}
#looking at overall abundance of species
BCI_sum <- sapply(BCI, sum)
boxplot(BCI_sum)

#What is the midpoint?
quantile(BCI_sum, 0.5)

BCI_total <- apply(BCI, 2, sum)

BCI_new <- rbind(BCI, BCI_total)

plot(density(log10(BCI_total)))
abline(v=log10(25), col='red')

#What is the midpoint?
quantile(BCI_sum, 0.5)

BCI_total <- apply(BCI, 2, sum)

BCI_new <- rbind(BCI, BCI_total)

#subset data to where everything below the midpoint is rare, and everything above is common
common = BCI
for (i in ncol(BCI_new)) {
  sp_abundance = sum(BCI[,1])
  if (sp_abundance < 25)
    common = common[,-i]
}

rare = BCI
for (i in ncol(BCI_new)) {
  sp_abundance = sum(BCI[,1])
  if (sp_abundance > 25)
    rare = rare[,-i]
}

```
```{r}
#is there spatial dependence in rare versus abundant species?
sr1 <- apply(common, 1, function(x) sum(x>0))
sr1_dist <- dist(sr1)
xy_dist <- dist(BCI_xy)
max_dist = max(xy_dist) / 2

common_mantel <- mantel(sr1_dist, xy_dist)
common_mantel

plot(xy_dist, sr1_dist)
abline(lm(sr1_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, sr1_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```


```{r}
sr2 <- apply(rare, 1, function(x) sum(x > 0))
sr2_dist <- dist(sr2)
xy_dist <- dist(BCI_xy)
max_dist = max(xy_dist) / 2

rare_mantel <- mantel(sr2_dist, xy_dist)
rare_mantel

plot(xy_dist, sr2_dist)
abline(lm(sr2_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, sr2_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)

```
Neither seem to differ from the null model. So there doesn't seem to be spatial dependence for abundant, nor rare, species.


2) Build two generalized linear models to predict the abundance of the species
*Drypetes standleyi* using the abundance of other tree species in the study site.
Specifically examine the following species as predictor variables:

```{r}
#Renaming specific species with more managable abbreviations
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")

sp_a <- BCI[,'Drypetes.standleyi']
sp_b <- BCI[, 'Cordia.lasiocalyx']
sp_c <- BCI[, 'Hirtella.triandra']
sp_d <- BCI[, 'Picramnia.latifolia']
sp_e <- BCI[, 'Quassia.amara']
sp_f <- BCI[, 'Tabernaemontana.arborea']
sp_g <- BCI[, 'Trattinnickia.aspera']
sp_h <- BCI[, 'Xylopia.macrantha']

BCI_short <- cbind(sp_a, sp_b, sp_c, sp_d, sp_e, sp_f, sp_g, sp_h)

```
Note renaming the species ids to something a little easier to work with like
"sp_a", "sp_b" will make model construction a little less cumbersome


* Model 1: only include a single species as a predictor variable
```{r}
#Model with Cordia.lasiocalyx
tree <- apply(BCI_short, 1, function(x) sum(x > 0))
tree_dat = data.frame(tree, BCI_short, BCI_xy)

tree_lm1 = gls(sp_a ~ sp_b, data=tree_dat)

par(mfrow=c(1,1))
plot(Variogram(tree_lm1, form= ~ x + y)) 
res = residuals(tree_lm1)
plot(dist(tree_dat[, c('x', 'y')]), dist(res))
lines(lowess(dist(tree_dat[, c('x', 'y')]), dist(res)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

tree_exp = update(tree_lm1, corr=corExp(form=~x + y)) 
plot(Variogram(tree_exp, maxDist = max_dist))

plot(Variogram(tree_exp, resType='normalized', maxDist = max_dist))

tree_rat_nug = update(tree_lm1, corr=corRatio(form=~x + y, nugget=T))

plot(Variogram(tree_rat_nug, maxDist = max_dist))
plot(Variogram(tree_rat_nug, resType='n', maxDist = max_dist))

anova(tree_lm1, tree_exp, tree_rat_nug, test=F)

summary(tree_lm1)
summary(tree_rat_nug)
```

* Model 2: include all of the species as predictor variables
```{r}
#Model with all species
tree_lm2 = gls(sp_a ~ sp_b + sp_c + sp_d + sp_e + sp_f + sp_g + sp_h, data=tree_dat)

par(mfrow=c(1,1))
plot(Variogram(tree_lm2, form= ~ x + y)) 
res = residuals(tree_lm2)
plot(dist(tree_dat[, c('x', 'y')]), dist(res))
lines(lowess(dist(tree_dat[, c('x', 'y')]), dist(res)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

tree_exp2 = update(tree_lm2, corr=corExp(form=~x + y)) 
plot(Variogram(tree_exp2, maxDist = max_dist))

plot(Variogram(tree_exp2, resType='normalized', maxDist = max_dist))

tree_rat_nug2 = update(tree_lm2, corr=corRatio(form=~x + y, nugget=T))

plot(Variogram(tree_rat_nug2, maxDist = max_dist))
plot(Variogram(tree_rat_nug2, resType='n', maxDist = max_dist))

anova(tree_lm2, tree_exp2, tree_rat_nug2, test=F)

summary(tree_lm2)
summary(tree_rat_nug2)
```

With both models examine the spatial dependence of the residuals using the 
function `Variogram`. Model the spatial dependence in the residuals using one
of the error structures available. 

* Did including the spatial error term have a large impact on the coefficients
of the model?
  Including the spatial error term in the GLS models made them quite a bit better in both cases. It also has an effect on the coefficients in the models.

* Did including the spatial error terms significantly improve model fit (use 
function `anova` to carry out model comparison)?
  When comparing the GLS with no spatial error to an exponential or the rational quadratic model, it does significantly improve using the error term. However, when comparing between the exponential and rational quadratic forms, there is no significant change in either case.

* Explain why you did or did not observe a difference in the influence of adding the spatial error term between the two models.
  By building the spatial error term into the model, we take into consideration the potential non-independence of samples that are closer together. Building that into the model makes the model better.