---
title: "Welling Univariate Modeling Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Univariate Assignment

Read in tree data, metadata can be found in: `./data/tree_metadata.txt`

```{r}
trees <- read.csv('treedata_subset.csv', header=T)

head(trees)
summary(trees)
```

1. Carry out an exploratory analysis using the tree dataset. 
Develop and compare models for species cover for a habitat generalist
[_*Acer rubrum*_ (Red maple)](http://www.durhamtownship.com/blog-archives/pix/November1407.jpg) 
and a habitat specialist [_*Abies fraseri*_ (Frasier fir)](https://upload.wikimedia.org/wikipedia/commons/d/d0/Abies_fraseri_Mitchell.jpg). 
Because this dataset includes both continuous and discrete explanatory
variables use the function `Anova` in the packages `car` as such.
This will estimate partial effect sizes, variance explained, and p-values for 
each explanatory variable included in the model.

```{r}
library(car)

#subseting the data 
A_rubrum <- trees[trees$species=="Acer rubrum",]

#graphing data
A_rubrum_sub<-subset.data.frame(A_rubrum, select = c(cover, elev, tci, streamdist, beers))
pairs(A_rubrum_sub, lower.panel = panel.smooth, upper.panel = panel.cor)
```

```{r}
#Exploratory analysis for A. rubrum:
Ar_mod1 <- lm(A_rubrum$cover~A_rubrum$elev+A_rubrum$tci+A_rubrum$streamdist+
                 A_rubrum$disturb+A_rubrum$beers)
summary(Ar_mod1)
Anova(Ar_mod1, type=3)

#based on initial model, disturbance and tci are not significant and will be 
# taken out of model
Ar_mod2 <- lm(A_rubrum$cover~A_rubrum$elev+A_rubrum$streamdist+
                 A_rubrum$beers)
summary(Ar_mod2)
Anova(Ar_mod2, type=3)


AIC(Ar_mod1)
AIC(Ar_mod2)
```
Without looking for interaction, it seems as though elevation, distance from streams, and beers
are important for the cover of this species, however the model with tci and disturbance built in is almost just as good.

```{r}
#Checking for interaction
Ar_mod3 <- lm(A_rubrum$cover~A_rubrum$elev*A_rubrum$tci*A_rubrum$streamdist*
                 A_rubrum$disturb*A_rubrum$beers)
Anova(Ar_mod3, type=3)
```
```{r}
plot(Ar_mod2)
```


There is not great evidence for a lot of interaction among variables. So, elevation, distance, and beers seems to explain A. rubrum cover best. 

```{r}
#subsetting the data
A_fraseri <- trees[trees$species=="Abies fraseri",]

#graphing data
A_fraseri_sub<-subset.data.frame(A_fraseri, select = c(cover, elev, tci, streamdist, beers))
pairs(A_fraseri_sub, lower.panel = panel.smooth, upper.panel = panel.cor)
```

```{r}
#Exploratory analysis for A. fraseri:
Af_mod1 <- lm(A_fraseri$cover~A_fraseri$elev+A_fraseri$tci+A_fraseri$streamdist+
                 A_fraseri$disturb+A_fraseri$beers)
summary(Af_mod1)
Anova(Af_mod1, type=3)

#based on the first model, it will only be run with elevation
Af_mod2 <- lm(A_fraseri$cover~A_fraseri$elev)
summary(Af_mod2)
Anova(Af_mod2, type=3)

AIC(Af_mod1)
AIC(Af_mod2)
```
Elevation seems to be the only important aspect for Abies fraseri cover.
However, taking out all other factors does not change the strength of the model
substantially.

```{r}
#Checking for interaction
Af_mod3 <- aov(A_fraseri$cover~A_fraseri$elev*A_fraseri$tci*A_fraseri$streamdist*
                 A_fraseri$disturb*A_fraseri$beers)
Anova(Af_mod3, type=3, singular.ok=TRUE)
```
There does not seem to be substantial evidence for interaction. 
```{r}
plot(Af_mod2)
```


Compare the p-values you observe using the function `Anova` to those generated
 using `summary`. 

For each species address the following additional questions:

How well does the exploratory model appear to explain cover?
  For A. rubrum, it takes multiple factors to explain cover, and the adjusted r-
  squared value is still low (0.03), so the model does not seem to explain cover well. 
  However, for A. fraseri, elevation helps explain much more about cover with an
  adjusted r-squared of 0.47. 
  
Which explanatory variables are the most important?
  For A. rubrum, elevation, distance from streams, and beers seem to be important,
  and for A. fraseri, only elevation seems important.
  
Do model diagnostics indicate any problems with violations of OLS assumptions?
  Neither of the best models found for the two species have normally distributed
  residuals.

Are you able to explain variance in one species better than another,
why might this be the case?
  The variance is more explained for A. fraseri. As a specialist, this species 
  will be in a specific type of habitat, and so environmental factors shouldn't be 
  as varied as for the generalist.


2. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

```{r}
Ar_glm1 <- glm(A_rubrum$cover~A_rubrum$elev+A_rubrum$tci+A_rubrum$streamdist+
                 A_rubrum$disturb+A_rubrum$beers, family='poisson')
Anova(Ar_glm1, type=3)

#based on above model, tci and disturbance will be taken out
Ar_glm2 <- glm(A_rubrum$cover~A_rubrum$elev+A_rubrum$streamdist+
                A_rubrum$beers, family='poisson')
Anova(Ar_glm2, type=3)

AIC(Ar_glm1)
AIC(Ar_glm2)
```


```{r}
Af_glm1 <- glm(A_fraseri$cover~A_fraseri$elev+A_fraseri$tci+A_fraseri$streamdist+
                 A_fraseri$disturb+A_fraseri$beers, family = 'poisson')
Anova(Af_mod1, type=3)

#based on the above model, it will only be run with elevation
Af_glm2 <- glm(A_fraseri$cover~A_fraseri$elev, family = 'poisson')
Anova(Af_glm2, type=3)

AIC(Af_glm1)
AIC(Af_glm2)
```


For assessing the degree of variation explained you can use a 
pseudo-R-squared statistic (note this is just one of many possible)

```{r}
pseudo_r2 = function(glm_mod) {
    1 -  glm_mod$deviance / glm_mod$null.deviance
}

pseudo_r2(Ar_glm2)
pseudo_r2(Af_glm2)
```

Compare the residual sums of squares between the traditional OLS 
 and glm models using `anova` (Note: not `Anova`) as such

```{r}
#for A. rubrum
anova(Ar_mod2, Ar_glm2)

#for A. fraseri
anova(Af_mod2, Af_glm2)
```

Does it appear that changing the error distribution changed the
results much? In what ways? 
Although the result does not seem largely changed, the residual sums of squares for 
the GLS models are much less than for OLS. But the important variables and variation
explained for each species has not changed.

3. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?
For A. rubrum, the variable that best help explain cover are elevation, distance from
streams, and the slope aspect, however, this model does not explain the variation 
well. For A. fraseri, elevation best explains cover and does so rather well compared to
the models built for A. rubrum.

4. (optional) Examine the behavior of the function `step()` using the 
exploratory models developed above. This is a very simple and not very
robust machine learning stepwise algorithm that uses AIC to select a 
best model. By default it does a backward selection routine. 
```{r}
step(Af_mod1, direction = "both")
```
```{r}
step(Ar_mod1, direction="both")
```



5. (optional) Develop a model for the number of species in each site 
(i.e., unique plotID). This variable will also be discrete so the Poisson
may be a good starting approximation. Side note: the Poisson
distribution converges asymptotically on the Gaussian distribution as the 
mean of the distribution increases. Thus Poisson regression does not differ
much from traditional OLS when means are large. 



