---
title: "Welling Multivariate modeling assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this assignment will be analyzing data on the Vegetation
and Environment in Dutch Dune Meadows.

To import the data and read the metadata run the following:

```{r}
library(vegan)
data(dune)
data(dune.env)
?dune

head(dune)
head(dune.env)
```

1. Conduct an indirect ordination on the dune plant community. Specifically,
visually examine a NMDS plot using the bray-curtis distance metric. Below is 
some code to help you develop a potential plot that emphasizes the role of the
environmental variable "Moisture". Describe how you interpret the
graphic. What is the goal of creating such a plot? Does this analysis suggest
any interesting findings with respect to the dune vegetation?

The sites that appear closest together have the higher degrees of similarity. This
type of plot is a good way to visualize groupings and/or clusters of similarities. It looks like the dune vegitation is arranged along a moisture gradient. 

```{r}
dune_mds <- metaMDS(dune)

plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
 # generate vector of colors 
color_vect = rev(terrain.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
           col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
           col=color_vect, pch=19)
    ```

2. Carry out a direct ordination using CCA in order to test any potential 
hypotheses that you developed after examining the MDS plot. Specifically,
carry out a test of the entire model (i.e., including all constrained axes)
and also carry out tests at the scale of individual explanatory variables
you included in your model if you included more than one variable. Plot your 
results.

plot(cca_tree, type='n', scaling=1)
orditorp(cca_tree, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_tree, display='bp', col='red')

```{r}
#all variables included
cca_dune <- cca(dune~., data=dune.env)
anova(cca_dune, by="margin", permutations=999)

plot(cca_dune, type='n', scaling=1)
orditorp(cca_dune, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune, display='bp', col='red')
```
After this analysis, only A1 and moisture seem to explain a large portion of variance.

```{r}
#only A1
cca_dune2 <- cca(dune~dune.env$A1)
anova(cca_dune2, by="margin", permutations=999)

plot(cca_dune2, type='n', scaling=1)
orditorp(cca_dune2, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune2, display='bp', col='red')
```
```{r}
#only moisture
cca_dune3 <- cca(dune~dune.env$Moisture)
anova(cca_dune3, by="margin", permutations=999)

plot(cca_dune3, type='n', scaling=1)
orditorp(cca_dune3, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune3, display='bp', col='red')
```

3. Do your two analyses agree with one another or complement one another or do
these two analyses seem to be suggesting different take home messages? Which
analysis do you find to be more useful?
